---
title: "Global GSEA Results"
author: "Hannah Damico"
date: '`r Sys.Date()`'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

options(bitmapType = "cairo", scipen = 9999)
library(tidyverse)
library(BiocManager)
library(statmod)
library(DataExplorer)
library(stringr)
library(ggprism)
library(plotly)
library(knitr)
library(DT)
library(readr)
library(limma)
library(reshape2)
library(pheatmap)
library(patchwork)
library(msigdbr)
library(readxl)
library(knitr)
library(kableExtra)
library(sgpv)
library(plotly)

# these have to be installed before you can use the imputeLCMD package
# if (!require("BiocManager", quietly = TRUE))
#     install.packages("BiocManager")
# 
# BiocManager::install(c("pcaMethods", "impute", "org.Mm.eg.db", "clusterProfiler", "pairwise_termsim"))

library(impute)
library(pcaMethods)
library(imputeLCMD)
library(clusterProfiler)
library(enrichplot)
library(org.Mm.eg.db)
library(DOSE)

top_proteins <- readRDS(here::here("Data/top_proteins.rds"))
set.seed(1234)
```



#  {.tabset}

Using second generation p-values, we observe a greater number of significant genes since we are not using such conservative methods as compared to BH FDR corrections. When using SGPVs, 0 indicates significant difference and 1 indicates equivalence. The following paper describes SGPVs more:  

[Second-generation p-values: improved rigor, reproducibility, & transparency in	
statistical	analyses](https://arxiv.org/pdf/1709.09333.pdf)


## gseGO & gseMKEGG {.tabset}
After adjusting for false positives, we found no gene enrichments. It is likely that we are not well powered enough to detect these enrichments. 

```{r}
df = top_proteins[!duplicated(top_proteins$Gene),]
df$logFC_T  <- -log10(df$P.Value)*sign(df$logFC) # this is Kin's preferred way of ordering genes

# transformed version
# we want the log2 fold change 
original_gene_listT <- df$logFC_T

# name the vector
names(original_gene_listT) <- df$Gene

# omit any NA values 
gene_listT<-na.omit(original_gene_listT)

# sort the list in decreasing order (required for clusterProfiler)
gene_listT = sort(gene_listT, decreasing = TRUE)

# normal logFC
# we want the log2 fold change 
original_gene_list <- df$logFC

# name the vector
names(original_gene_list) <- df$Gene

# omit any NA values 
gene_list<-na.omit(original_gene_list)

# sort the list in decreasing order (required for clusterProfiler)
gene_list = sort(gene_list, decreasing = TRUE)



```

### gseGO Dotplot

Ontology =  BP, CC, MF  

```{r eval = T, warning = FALSE, message=FALSE, fig.height=20, fig.width=12}
# other method
# reading in data from deseq2

set.seed(1234)

gse <- gseGO(geneList=gene_listT, 
             ont ="ALL", 
             keyType = "SYMBOL", 
             minGSSize = 3, 
             maxGSSize = 800, 
             pvalueCutoff = 0.05, 
             eps = 0, 
             verbose = TRUE, 
             OrgDb = org.Mm.eg.db::org.Mm.eg.db, 
             pAdjustMethod = "BH") # keep BH to adjust for false positives

symbols <- mapIds(org.Mm.eg.db::org.Mm.eg.db, keys = names(gene_listT), keytype = "SYMBOL", column="ENTREZID")

# gene_kegg <- gene_list
# names(gene_kegg) <- symbols
# 
# gse_KEGG <- gseMKEGG(geneList=na.omit(gene_kegg), 
#              organism = "mmu",
#              keyType = "ncbi-geneid", 
#              minGSSize = 3, 
#              maxGSSize = 800, 
#              pvalueCutoff = 0.05, 
#              verbose = TRUE,
#              pAdjustMethod = "BH") 

# ont = biological processes
# celullar component, MF = molecular function - 


require(DOSE)
dotplot(gse, showCategory=500, split=".sign") + facet_grid(.~.sign)

sv <- dotplot(gse, showCategory=1000, split=".sign")[1]


# save gseGO object
saveRDS(gse, "../Data/gseGO.rds")
```

```{r}
data.frame(gse$Description, gse$ONTOLOGY) %>% filter(gse.Description == "actin catabolic membrane interferon-alpha")
```


### GO Enrichment Scores

```{r 'enrichement_scores', cache = T, results='asis'}
enrichment_scores <- data.frame(ID = gse$ID, EnrichmentScore = gse$enrichmentScore) %>%  
           mutate(Direction = factor(case_when(EnrichmentScore > 0 & abs(EnrichmentScore) > 0 ~ "Activated in KOs",
                                 EnrichmentScore < 0 & abs(EnrichmentScore) > 0 ~ "Surpressed in KOs"), 
                                 levels = c("Surpressed in KOs", "Activated in KOs")))

enrichment_scores <- enrichment_scores %>% group_by(Direction) %>% arrange(EnrichmentScore)

table(enrichment_scores$Direction)


enrichment_scores %>% kbl(full_width = F)  %>% kable_styling() %>% pack_rows("Surpressed in KOs", 1,54) %>% pack_rows("Activated in KOs", 55, nrow(enrichment_scores)) %>% scroll_box(height = "600px", width = "800px")
```


### Enriched GO Pathways

```{r results='asis'}
library(data.table)
ME <- lapply(gse$core_enrichment, function(x){unlist(strsplit(gsub("*/", " ", x), " "))})
names(ME) <- gse$ID


## work as of 10/01
core_genes <- lapply(gse$core_enrichment, function(x){
  genes <- unlist(strsplit(gsub("*/", " ", x), " "))
  df <- data.frame(genes)
  return(df)
  })

names(core_genes) <- gse$Description

core_genes_df <- rbindlist(core_genes, idcol="pathway")

NES <- lapply(c(gse$NES), function(NES){
  df <- data.frame(NES = NES)
return(df)
})
names(NES) <- gse$Description
NES_df <- rbindlist(NES, idcol = "pathway") 

pvalue <- lapply(c(gse$pvalue), function(pvalue){
  df <- data.frame(pvalue = pvalue)
return(df)
})
names(pvalue) <- gse$Description
pvalue_df <- rbindlist(pvalue, idcol = "pathway") 


p.adjust <- lapply(c(gse$p.adjust), function(padj){
  df <- data.frame(p.adjust = padj)
return(df)
})
names(p.adjust) <- gse$Description
p.adjust_df <- rbindlist(p.adjust, idcol = "pathway") 

goIDs <- lapply(c(gse$ID), function(ID){
  df <- data.frame(ID = ID)
return(df)
})
names(goIDs) <- gse$Description
ID_df <- rbindlist(goIDs, idcol = "pathway") 



GO_df <- merge(merge(pvalue_df, p.adjust_df, by = c("pathway")),merge(NES_df, core_genes_df, by = "pathway"), by = "pathway") %>% relocate(genes)

GO_df_final <- merge(GO_df, ID_df, by = "pathway") %>% relocate(ID, pathway, genes, NES)
# write.csv(GO_df_final, "Data-20241001//GO_pathway_genes_20241001.csv", row.names = F)
```

```{r}

#lapply(ME, head)
your_list <- lapply(ME, function(x) purrr::map_chr(x, as.character))

# Convert list to data frame
df <- data.frame(
  Element = names(your_list),
  Items = I(your_list),
  stringsAsFactors = FALSE
)

df_ME <- data.frame(do.call(cbind, ME))

df_ME_melted <- pivot_longer(df_ME, cols = colnames(df_ME), names_to = "Pathway", values_to = "Genes")
# saveRDS(df_ME_melted, "../Data/df_ME_melted.rds")
# df_ME_melted  %>% group_by(Pathway) %>% count
# a file that summarizes the genes that make up each GO term/pathway in the averaged NES analysis
# df_ME %>% kbl(caption = "Enriched GO Pathways") %>% kable_styling()  %>% scroll_box(width = "800px", height = "600px")
df_ME_melted  %>% group_by(Pathway) %>% arrange(Pathway)
```


```{r 'enriched_go_pathways', cache = T, }
library(Hmisc)
EP <- data.frame(ID = gse$ID, Description = gse$Description, adj.P.Value = gse$p.adjust)
EP %>% filter(adj.P.Value < 0.0005) %>% kbl(caption = "Enriched GO Pathways") %>% kable_styling()  %>% scroll_box(width = "800px", height = "600px")

# Hmisc::label(sv$data$Description) <- sv$data$ID

#  %>% rename("raw_pvalue" = "pvalue") 
sv_tab <- sv$data %>% dplyr::select(ID, Description,pvalue, p.adjust, enrichmentScore)%>% kbl(caption = "Enriched GO Pathways") %>% kable_styling()  %>% scroll_box(width = "800px", height = "900px")

sv_data <- sv$data


# saveRDS(sv_data, file = "sv_data.rds")
# write.csv(sv$data[!colnames(sv$data) %in% c("rank", "leading_edge", "core_enrichment", "x")], file = "all_enriched_pathways.csv")
```



```{r}
ME <- lapply(gse$core_enrichment[gse$ID %in% sv$data$ID], function(x){unlist(strsplit(gsub("*/", " ", x), " "))})

names(ME) <- gse$Description[gse$ID %in% sv$data$ID]

#lapply(ME, head)
df_ME <- data.frame(do.call(cbind, ME))
# create RDS of this!
saveRDS(df_ME, "../Data/df_ME.rds")

df_ME_melted <- pivot_longer(df_ME, cols = colnames(df_ME), names_to = "Pathway", values_to = "Genes")
# df_ME_melted  %>% group_by(Pathway)

gene_tab <- df_ME %>% kbl(caption = "Enriched GO Pathways") %>% kable_styling()  %>% scroll_box(width = "1000px", height = "600px")
gene_tab2 <- df_ME
# saveRDS(gene_tab, file = "gene_tab.rds")
write.csv(gene_tab2, "../Data/all_enriched_pathway_genes_20241001.csv")
```



```{r 'dotplot_all', cache = T, fig.height=65, fig.width=15, full_width = T }
set.seed(1234)
dp_all <- dotplot(gse, showCategory=100, split=".sign")
dp_all + ggtitle("All Pathways - gseGO")


# saveRDS(dp_all, "dp_all.rds")
# ggsave("dp_all.jpeg", dp_all, height = 65, width = 15, limitsize = FALSE)
```


```{r 'dotplot_0.0004', cache = T, fig.height=18, fig.width=16, full_width = T}


library("tidytree")
library("treeio")
#library("ggtreeExtra")
# sessionInfo()
# ggtreeExtra_1.12.0     treeio_1.26.0          tidytree_0.4.6   
set.seed(1234)
edox2 <- pairwise_termsim(gse)
p1 <- treeplot(edox2, cluster.params = list(n = 6)) + ggtitle("Highly significant genes - p < 0.0004") # default clustering method is ward.D
# saveRDS(p1, "tree_wardD.rds")
p2 <- treeplot(edox2, cluster.params = list(method = "average", n = 6))
# saveRDS(p2, "tree_avg.rds")
# aplot::plot_list(p1, tag_levels='A')

```






```{r 'dotplot_0.05', cache = T, fig.height=20, fig.width=16, full_width = T}
# dendrogram showing p < 0.05
set.seed(1234)
tp.0.05 <- treeplot(edox2, showCategory = 250, cluster.params = list(n = 9)) + ggtitle("Significant genes - p < 0.05") 
tp.0.05
# saveRDS(tp.0.05, "tp.0.05.rds")

```


```{r 'dotplot_0.01', cache = T, fig.height=18, fig.width=16, full_width = T}
# dendrogram showing p < 0.01
set.seed(1234)
tp.0.01 <- treeplot(edox2, showCategory = 80,cluster.params = list(n = 9)) + ggtitle("Significant genes - p < 0.01") 
tp.0.01
# View(tp.0.01)
# data.frame(tp.0.01$data) %>% group_by(group) %>% count()
# grab cluster labels! order by y so they're in the correct order
labs_clus <- ggplot_build((tp.0.01))$data[[3]] %>% arrange(desc(y))
labs_clus <- labs_clus$label
labs_clus <- gsub("\n", " ", labs_clus) # remove \n new line 
```




```{r }
test <- data.frame(tp.0.01$data)
# test <- test[order(test$y),]
test <- na.omit(test)
# unique(test$group)

cluster_order <- test %>% arrange(desc(y))
cluster_order <- unique(cluster_order$group)
# weighted avg. of enrichment scores
# saveRDS(tp.0.01, "tp.0.01.rds")
# enrichment on the union of genes in sets

paste(labs_clus, cluster_order)

```


```{r}
# grab relevant columns
cluster_genes <- function(cluster_name){
  # grab label names from from cluster 1 and match to column names in df_ME that contains the genes
  cluster_labels <- test$label[(test$group == cluster_name) & test$isTip == TRUE]
  # need slight adjustments, add . instead of space, now they should match colnames of df_ME
  cluster_labels <- gsub(" ", ".", cluster_labels)
  # select labels for cluster 1 from df_ME
  cluster_genes <- df_ME[colnames(df_ME) %in% cluster_labels]
  # add intersection check 
  
  return(cluster_genes)
    
}


# these are the cluster levels
cls <- lapply(cluster_order, cluster_genes)
names(cls) <- labs_clus

find_union <- function(df) {
  Reduce(union, df) # union, not intersection 
}

# avg. 

# Apply the function to each dataframe in the list
union_results <- lapply(cls, find_union)
# now take these intersecting genes and perform a new enrichment as if they were custom lists


# saveRDS(union_results, "union_results.rds")
```


function takes names of groups from group variable
for every cluster, switch a space to a .
check matching genes
print out intersecting gene list


```{r 'enrichment_score_plot', cache = T, fig.width=30, fig.height=30, full_width = T}
temp <- sv$data
ord <- temp %>% group_by(Description) %>% arrange(desc(enrichmentScore))
temp$Description <- factor(temp$Description, levels = ord$Description)

enrichment_score_plot <- temp %>% 
  ggplot() + 
  geom_point(aes(Description, enrichmentScore, color  = p.adjust)) + coord_flip() + ggtitle("Top Gene Enrichment Scores") + theme(plot.caption = element_text(hjust = 0)) + labs("Note that adj. p-value comes from gseaGO results.") + facet_grid(.~`.sign`)

# saveRDS(enrichment_score_plot, "enrichment_score_plot.rds")
ESP_plotly <- ggplotly(enrichment_score_plot, height = 1700, width = 1500)

# saveRDS(ESP_plotly, "ESP_plotly.rds")
```



&nbsp;
&nbsp;
&nbsp;

### Session Info

```{r}
sessionInfo()
```


a file that summarizes the genes that make up each GO term/pathway in the averaged NES analysis
